Seismic parameter estimation and the Canadian crust

### 2013-09-20 paper/draftBP.tex
interpolated stack plot

Mean crustal composition is an important parameter characterizing continental crust that should not be confused with actual lithological or mineralogical state
of the crust. Rather, mean crustal composition is a representative quantity that serves as a constraint on crustal structure and evolution in space and time. Early
attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks [Rudnick and Gao, 2003]. After the plate-tectonic revolution, Taylor [1967, 1977], proposed the 'island arc' or 'andesite' model which posited that bulk crustal composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, island arcs are more basaltic than andesitic and their accretions will lead to basaltic crustal growth and therefore cannot account for the intermediate composition of post-Archean crust. The recognition of these shortcomings led to the development of seismic methods to interrogate middle and lower crustal levels and place constraints on their composition [Christensen and Mooney, 1995; Rudnick and Fountain, 1995]. Unlike these latter studies, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. [1981] that combines $V_P/V_S$ from teleseismic measurements with contraints on $V_P$ from previous active source studies as presented in the crust 1.0 model.

@@

In Figure \ref{fig:lith} we plot our area-weighted seismic parameter estimates for the entire Canadian data set along with area-weighted average estimates for various tectonic subdivisions on a plot $V_P$ versus $V_S$. Also plotted are values for a selection of mineral assemblages considered to be representative of common crustal rocks from the compilation of laboratory measurements by Christensen [1996]. Dashed lines delimit isocontours in $V_P/V_S$. The choice of common shield lithologies is motivated by observations along 30km of paleo-depth represented in the Kapuskasing uplift of the Superior province [Percival and West, 1994]. The Kapuskasing uplift comprises Archean granite-greenstone terranes dominated by granitoid and gneissic rocks, and contains supracrustal assemblages generally metamorphosed from greenschist to amphibolite facies (hence the inclusion of hornblendite and amphibolite) and early plutonic structures comprised largely of tonalites. Tonalites and granodiorites are also part of the TTG (tonalite-trondhjemite-granodiorite) suite which dominates Archaean cratons [Rudnick, 1995]. Seismic velocity values for these lithologies are taken from the work of Christensen [1996] and are quoted at average crustal pressures of 600 MPa.

The Slave, Superior and Churchill provinces share similar mean crustal compositions, with the Slave province being slightly more tonalitic in average composition. These three provinces cluster around the average for the Canadian Shield. The Grenville Province on the other hand differs significantly in mean composition. One interpretation is that the crust in the Grenville province has a mean crustal composition corresponding to a more mafic lower crustal composition than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the South-East of the province also have seismic ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). The average $V_P/V_S$ value for the Grenville Province is 1.77 marginally higher than the average value computed by the study of Eaton et. al. [2005].

We now take the

[ Continue with crustal composition discussion ]

Other researchers have focused on Poisson's Ratio [Zandt and Ammon, 1995] as an indicator of bulk crustal composition as well as crustal thickness and P-wave velocity [Durrheim and Mooney, 1991] to constrain models of crustal formation. Zandt and Ammon [1995] report high $V_P/V_S$ values for Archean Shields and conclude that no systematic difference between Archaean and Proterozoic crust exists. Their average, 1.84, much higher than the wieghted average for the Canadian Shield of 1.74 computed herein. The Authors also report a low $V_P/V_S$ value for Cenozoic-Mesozoic crust, 1.73 compared to the value calculated here of 1.77. Zandt and Ammon interpret these results showing high values for older crust and lower values for younger crust as indicating two end member scenarios. In the uniformitarian model, a delamination type process operates to remove mafic lower crust during continental collisions and is subsequently followed by a stabilizing process involving basaltic underplating. This implies that younger crust has not yet completed the crustal genesis cycle. In their second scenario, Precambrian cratons evolve through unique processes and younger orogens involve recycling and reworking existing material with little crustal growth.

Whereas Zandt and Ammon show increasing $V_P/V_S$ values with crustal age the estimates computed in this study indicate an opposite trend, that is lower seismic velocity ratios with crustal age. This aligns more closely to the results of Durrheim and Mooney [1991] who provide results showing a thick higher velocity Proterozoic crust compared to a thinner lower velocity Archean crust. The crustal thickness data attained in this study also show thicker crusts for Proterozoic units (38.9 km) than Archean units (38.0 km) though this difference is not as drastic as the greater than 5km difference reported by Durrheim and Mooney. Though $Vp$ data was not explicity estimated in this study, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. Though Durrheim and Mooney accept the hypothesis that Proterozoic crust is more prone to basaltic underplating, thereby explaining higher velocities and thicker crust, Roberta Rudnick [1995; Rucknick and Gao, 2003] has argued persuasively that a range of possible processes may explain this data.

[ Use crustal composition plus increasing Vp/Vs with younger crust result to make ``best'' pick among the 4 main processes Rudnick proposes to complete section.]

>>>>

Seismic reflectivity of the continental crust is a valuable constraint on the processes governing crustal evolution. Research into the Mohoveric discontinuity and crustal thickness is ongoing and providing new insights into ...

1. introductory paragraph along the lines of the one you did for crustal composition, identifying the
importance of 1-D crustal structure for models of evolution. Introduce Conrad discontinuity and
general seismic reflection subdivision of crust into more transparent upper half and reflective
lower half, ie, summarize basic ideas from Brown EOS article on Conrad and Brocher and Mooney
GJRAS article

2. cite station sorted figure + perhaps normalized stack figure (as per above) and discuss where
energy lies, most pronounced near surface and Moho, lower energy levels in bottom 1/3

3. draw association between bottom 1/3 of low teleseismic reflectivity and high lower crustal  (seismic
reflection) reflectivity, explain effect of frequency in different response

4. discuss evidence for Conrad as distinct discontinuity (doesn't appear to be) versus the more
general subdivision of upper and lower crusts with distinct properties.

Mean crustal composition is an important parameter characterizing continental crust that should not be confused with actual lithological or mineralogical state of the crust. Rather, mean crustal composition is a representative average quantity that serves as a constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks [Rudnick and Gao, 2003]. After the plate-tectonic revolution, Taylor [1967, 1977], proposed the 'island arc' or 'andesite' model which posited that bulk crustal composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition [Christensen and Mooney, 1995; Rudnick and Fountain, 1995]. Unlike these latter studies, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. [1981] that combines $V_P/V_S$ from teleseismic measurements with contraints on $V_P$ from previous active source studies as presented in the crust 1.0 model.

@@

In Figure \ref{fig:lith} we plot our area-weighted seismic parameter estimates for the entire Canadian data set along with area-weighted average estimates for various tectonic subdivisions on a plot $V_P$ versus $V_S$. Also plotted are values for a selection of mineral assemblages considered to be representative of common crustal rocks from the compilation of laboratory measurements by Christensen [1996], all quoted at average crustal pressures of 600 MPa. Dashed lines delimit isocontours in $V_P/V_S$. The choice of common shield lithologies is motivated by observations along 30km of paleo-depth represented in the Kapuskasing uplift of the Superior province [Percival and West, 1994]. The Kapuskasing uplift comprises Archean granite-greenstone terranes dominated by granitoid and gneissic rocks, and contains supracrustal assemblages generally metamorphosed from greenschist to amphibolite facies (hence the inclusion of hornblendite and amphibolite) and early plutonic structures comprised largely of tonalites. Tonalites and granodiorites are also part of the TTG (tonalite-trondhjemite-granodiorite) suite which dominates Archaean cratons [Rudnick, 1995].

Seismic parameter estimates for Canada and the Canadian Shield fall, to a first-order approximation, between those of two candidate end-member lithologies: granite geniss and diorite. This intermediate or andesitic composition is in general agreement with other earlier studies of bulk crustal composition [Rudnick and Fountain, 1995; Christensen and Mooney [1995]. The Slave, Superior and Churchill provinces possess comparable seismic parameters implying similar mean crustal compositions, with the Slave province represented as slightly more tonalitic in average composition. All three provinces cluster about the average value for the Canadian Shield. The Grenville Province, in contrast, displays markedly lower $V_S$ and higher $V_P/V_S$ implying a significant difference in mean composition. The high $V_P/V_S$ estimate, 1.77, is similar to the averaged results from a previous seismic study [Eaton et. al., 2005] that included stations of the central Grenville Orogen, and might be interpreted to indicate that the Grenville province possesses a mean crustal composition corresponding composition that is more mafic than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the south-east of the province also display $V_P/V_S$ ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). Moreover, and more importantly, the average $V_P$ values for the Grenville province (a region well sampled by refraction profiles) requires low $V_S$ values that are inconsistent with a more mafic composition. Alternative explanations that allow for both high $V_P/V_S$ and reduced $V_P$, $V_S$ include the possibility that Grenville crust contains higher fluid concentrations than that of other tectonic provinces. (NIK?? any others?)

\begin{table}
  \begin{tabular}{ p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} p{2cm} }
    \hline
    Oxide, wt\% & Smithson [1978] & Rudnick and Fountain [1995] & Christensen and Mooney [1995] & Rudnick and Gao [2003] & This Work \\
    \hline
    $SiO_2$   & 63.0 & 52.6 & 61.7 & 60.6 & 64.3 \\
    $TiO_2$   & 0.7 & 0.8 & 0.9 & 0.7 & 0.6 \\
    $Al_2O_3$ & 15.8 & 16.6 & 14.7 & 15.9 & 15.5 \\
    $Fe_2O_3$ & 2.0 & --- & 1.9 & --- & 1.9 \\
    $FeO$     & 3.4 & 6.6 & 5.1 & 6.7 & 3.3 \\
    $MnO$     & 0.1 & 0.1 & 0.1 & 0.1 & 0.1 \\
    $MgO$     & 2.8 & 4.4 & 3.1 & 4.7 & 2.2 \\
    $CaO$     & 4.6 & 6.4 & 5.7 & 6.4 & 4.3 \\
    $Na_2O$   & 4.0 & 3.2 & 3.6 & 3.1 & 3.6 \\
    $K_2O$    & 2.7 & 1.9 & 2.1 & 1.8 & 2.9 \\
    $H_2O$    & --- & --- & 0.8 & --- & 1.1 \\
    $P_2O_5$  & --- & 0.2 & 0.2 & 0.1 & 0.2 \\
    $CO_2$    & --- & --- & --- & --- & 0.1 \\
    \hline
  \end{tabular}
  \caption{Bulk crustal composition comparison from selected sources. Rudnick and Fountain [1995] and Rudnick and Gao [2003] use a combined $Fe$ quantity and is listed under $FeO$}
\label{table:composition}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{composition}
  \caption{Comparison of major crustal elements from different studies. All estimates have been normalized to the results computed in this study.}
  \label{fig:composition}
\end{figure}

Since our average seismic parameter estimates for the entire Canadian crust lie on a line connecting diorite with granite gneiss (\ref{fig:lith}), we may estimate bulk crustal composition in proportion to these lithologies as 51\% diorite and 49\% granite. The corresponding chemical breakdown is provided in table \ref{table:composition} along with previous estimates from global crustal studies. The chemical estimates produced in our analysis reveal a crust with marginally higher average $SiO_2$ composition and lower $Fe$ and $Mg$ content than these previous studies. It is noteworthy that that the two studies involving greatest reliance on seismic attributes, ours and the {\it P}-wave velocity study by Smithson [1978], bear the closest similarity. Figure \ref{fig:composition} displays the major crustal element contributions from the data in table \ref{table:composition} normalized to our average Canadian crustal composition. The gray-shaded region indicates the $\pm 30\%$ range from our reference average Canadian crustal model.  Also included in this figure is the our average crustal composition estimate for the Canadian Shield. Its values closely track the global crustal estimates of Christensen \& Mooney [1995] and Smithson [1978]. These data imply that the extensive platform and orogenic regions of Canada skew the average composition away from the global estimates of previous studies through incorporation of a higher proportion of granite gneiss. In contrast, the Canadian Shield crust aligns more closely with these previous global estimates and is represented as 68\% dioritic and 32\% granitic using the two end-member model.

Other researchers have employed Poisson's ratio [Zandt and Ammon, 1995] alone as an indicator of bulk crustal composition, and crustal thickness and {\it P}-wave velocity [Durrheim and Mooney, 1991] to constrain models of crustal formation. Zandt and Ammon [1995] report high $V_P/V_S$ values for Archean Shields and concluded that no systematic difference between Archean and Proterozoic crust exists. Their average, $V_P/V_S=1.84$, is much higher than our weighted average for the Canadian Shield of 1.74. These authors also report a low $V_P/V_S$ value for Cenozoic-Mesozoic crust, 1.73 compared to the corresponding value for our data of 1.77. They interpret these high values for older crust and low values for younger crust as favouring either of two limiting models. In the uniformitarian model, a delamination type process operates to remove mafic lower crust during continental collisions and is subsequently followed by a stabilizing process involving basaltic underplating. This model implies that younger crust has yet to complete the full crustal genesis cycle. In their second model, Precambrian cratons evolve through unique processes and younger orogens involve the recycling and reworking of existing material with little net crustal growth.

Whereas Zandt and Ammon present evidence for increasing $V_P/V_S$ with crustal age, our estimates signify the opposite trend. Our results are more consistent with those of Durrheim and Mooney [1991] who argue that Proterozoic crust is thicker with higher {\it P}-velocity than Archean crust. The crustal thickness data presented in this study is only marginally thicker for Proterozoic crust (38.9 km) than Archean crust (38.0 km) and this difference is significantly smaller than the 5 km difference reported by Durrheim and Mooney. Though the difference is not large, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. The hypothesis that Proterozoic crust is prone to basaltic underplating, thereby explaining higher velocities and thicker crust, suggested by Durrheim and Mooney [1991], has been persuasively augmented by recent work [Rudnick 1995; Rucknick and Gao, 2003] to now include a range of possible processes that may explain the data.

One possible process is the preferential recycling of soluble ions such as $Mg^{2+}$ and $Ca^{2+}$. These ions are weathered and carried into the oceans leaving behind silicon and aluminum [Rudnick and Gao, 2003] producing a net change to crustal composition over time. In another model, deep crustal roots expose lower crust to processes leading to eclogitization and subsequent delamination due to a density inversion with the underlying mantle peridoties. This model does not directly address the question as to why there is an apparent variance in average velocities between younger and older crusts, unless the delamination process has evolved over time. However, knowledge of the spatial or temporal extent of delamination is difficult to document [Rudnick, 1995]. A third model suggests that a mixture of silicic melts from subducting oceanic crust and basaltic melts from mantle peridotites were important in early crustal formation. This process is thought to be more common in the hotter Archean and is related to the TGG suites of Archean granitoids [Rudnick and Gao, 2003] and therefore provides a possible explanation for the data. It is important to state that all of these processes, as well as others, may have been operating throughout geologic time.

This study provides results that are in general agreement with other recent crustal studies and similarly argues for an intermediate or andesitic crust. There are several processes that have been suggested to explain this intermediate composition ...
